### 2016-05-04 text/Chapter06.tex
Add structural coeffect proofs (Chapter 6)

\begin{equation*}
\begin{array}{l}
\tyrule{counit}
  {\Gamma \vdash e : \ctyp{\alift{\cunit}}{\tau}}
  {\Gamma \vdash \ident{counit}_{\cunit}~e : \tau} \\
\\
\tyrule{cobind}
  {\Gamma \vdash e_1 : \ctyp{\aclrd{\textbf{r}}}{(\tau_1 \xtimes \ldots \xtimes \tau_k)} \rightarrow \tau & \Gamma \vdash e_2 : \ctyp{\cclrd{s}\aseq\aclrd{\textbf{r}}}{\tau_1(\tau_1 \xtimes \ldots \xtimes \tau_k)} }
  {\Gamma \vdash \ident{cobind}_{\cclrd{r}, \cclrd{s}}~e_1~e_2 : \ctyp{\alift{\cclrd{s}}}{\tau}} \\
  \\
\tyrule{merge}
  {\Gamma \vdash  e : \ctyp{\aclrd{\textbf{r}}}{(\tau_1 \xtimes \ldots \xtimes \tau_l)} \times \ctyp{\aclrd{\textbf{s}}}{(\tau_{l+1} \xtimes \ldots \xtimes \tau_k)} }
  {\Gamma \vdash  \ident{merge}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}}~e : \ctyp{\aclrd{\textbf{r}}\atimes\aclrd{\textbf{s}}}{(\tau_1 \xtimes \ldots \xtimes \tau_k)} }\\
\\
\tyrule{split}
  {\Gamma \vdash  e : \ctyp{\aclrd{\textbf{r}}\atimes\aclrd{\textbf{s}}}{(\tau_1 \xtimes \ldots \xtimes \tau_k)} }
  {\Gamma \vdash  \ident{split}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}}~e : \ctyp{\aclrd{\textbf{r}}}{(\tau_1 \xtimes \ldots \xtimes \tau_l)} \times \ctyp{\aclrd{\textbf{s}}}{(\tau_{l+1} \xtimes \ldots \xtimes \tau_k)}} \\
\\
\tyrule{dup}
  {\Gamma \vdash  e : \ctyp{\alift{\cclrd{r}\cpar\cclrd{s}}}{\tau} }
  {\Gamma \vdash  \ident{dup}_{\cclrd{r},\cclrd{s}}~e : \ctyp{\alift{\cclrd{r},\cclrd{s}}}{(\tau \xtimes \tau)} } \\
\end{array}
\end{equation*}

@@

\begin{theorem}[Well-typedness of the translation]
\label{thm:semantics-welltyped}
Given a typing derivation for a well-typed closed expression $\coctx{}{\alift{}} \vdash e : \tau$
written in a structural context-aware programming langugae that is translated to the target language as
(we write $\ldots$ for the omitted part of the translation tree):
%
\begin{equation*}
  \semdeff
    {~~(\ldots)~~}
    {\coctx{}{\alift{}} \vdash e : \tau }
    {(\ldots)}
    { ~~f }
\end{equation*}
%
Then $f$ is well-typed, \ie~in the target language: $\vdash f : \sem{\coctx{()}{\alift{}}} \rightarrow \sem{\tau}$.
\end{theorem}
\begin{proof}
By rule induction over the derivation of the translation. Given a judgement
$\coctx{x_1\!:\!\tau_1 \ldots x_n\!:\!\tau_n}{\aclrd{\textbf{c}}} \vdash e : \tau$ where
$\aclrd{\textbf{c}} = \alift{\cclrd{c_1},\ldots,\cclrd{c_n}}$, the translation
constructs a function of type $\ctyp{\aclrd{\textbf{c}}}(\sem{\tau_1}\xtimes\ldots\xtimes\sem{\tau_n}}) \rightarrow \sem{\tau}$.

@@

%
%
%
%
TODO
%
%
%
%
%

\vspace{0.25em}\noindent\hangindent=0.6cm
Case (\emph{abs}): The type of $\ctx$ is $\ctyp{\cclrd{r}}(\ldots)$ and the type of $v$ is $\ctyp{\cclrd{s}}\tau_1$,
  calling $\ident{merge}_{\cclrd{r},\cclrd{s}}$ and reassociating produces
  $\ctyp{\cclrd{r}\czip\cclrd{s}}(\ldots)$ as expected by $f$.

\vspace{0.25em}\noindent\hangindent=0.6cm
Case (\emph{app}): After applying $\ident{split}_{\cclrd{r}, \cclrd{s}\cseq\cclrd{t}}$, the types of
  $\ctx_1, \ctx_2$ are $\ctyp{\cclrd{r}}(\ldots)$ and $\ctyp{\cclrd{s}\cseq\cclrd{t}}(\ldots)$, respectively.
  $g$ requires $\ctyp{\cclrd{s}}(\ldots)$ and so the result of $\ident{cobind}_{\cclrd{s},\cclrd{t}}$ is
  $\ctyp{\cclrd{t}}\tau_1$ as required by $f$.
\end{proof}

@@

\subsection{Safety of concrete thing}
\label{sec:struct-transl-safety}

\newpage

@@

This chapter completes the key development of this thesis -- the presentation of the coeffect
framework, consisting of two calculi capturing properties of context-aware computations introduced
in Chapter~\ref{ch:applications}. In the previous chapter, we focused on the whole-context
properties of computations and we developed \emph{flat coeffect calculus} to capture them.
This chapter develops \emph{structural coeffect calculus}, which captures \emph{per-variable}
contextual properties. The system provides a precise analysis of liveness and dataflow and allows
other interesting uses such as tracking of variable accesses based on bounded linear logic.

Following the structure of the previous chapter, the structural coeffect calculus is parameterized
by a \emph{structural coeffect algebra}. The two definitions are similar -- both require operations
$\cseq$ and $\cpar$ that model sequential and pointwise composition, respectively. For flat
coeffects, we required $\czip$ to model context merging. For structural coeffects, we
instead use a vector (free monoid) with the $\atimes$ operation -- which serves a similar purpose
as $\czip$.

In order to keep track of separate annotations for each variable, we use a system with
explicit structural rules (contraction, weakening and exchange) that manipulate the structure
of variables and the structure of annotations at the same time.

The chapter presents two technical results. The first is the semantics of structural coeffect
calculus in terms of \emph{structural indexed comonads}. The semantics serves two purposes --
it unifies our three examples (bounded reuse, dataflow and liveness) and it demonstrates
suitability of the type system. The second technical result is equational theory for the structural
coeffect calculus. In particular, we show that $\beta$-reduction and $\eta$-expansion have the
typing preservation property for any instance of the structural coeffect calculus. These two
strong properties are desirable for programming languages, but are often not satisfied
(\eg~by languages with effects).

Two questions remain open for the upcoming chapter. Firstly, is there a way to present
flat and structural coeffect calculi in a uniform way as a single system? Secondly, what
is the best pathway towards a practical implementation of programming languages based on
coeffect systems?

>>>>

\begin{equation*}
\begin{array}{l}
\tyrule{counit}
  {\Gamma \vdash e : \ctyp{\alift{\cunit}}{\tau}}
  {\Gamma \vdash \ident{counit}_{\cunit}~e : \tau} \\
\\
\tyrule{cobind}
  {\Gamma \vdash e_1 : \ctyp{\aclrd{\textbf{r}}}{(\tau_1 \xtimes \ldots \xtimes \tau_k)} \rightarrow \tau & \Gamma \vdash e_2 : \ctyp{\cclrd{s}\aseq\aclrd{\textbf{r}}}{\tau_1(\tau_1 \xtimes \ldots \xtimes \tau_k)} }
  {\Gamma \vdash \ident{cobind}_{\cclrd{s}, \aclrd{\textbf{r}}}~e_1~e_2 : \ctyp{\alift{\cclrd{s}}}{\tau}} \\
  \\
\tyrule{merge}
  {\Gamma \vdash  e : \ctyp{\aclrd{\textbf{r}}}{(\tau_1 \xtimes \ldots \xtimes \tau_l)} \times \ctyp{\aclrd{\textbf{s}}}{(\tau_{l+1} \xtimes \ldots \xtimes \tau_k)} }
  {\Gamma \vdash  \ident{merge}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}}~e : \ctyp{\aclrd{\textbf{r}}\atimes\aclrd{\textbf{s}}}{(\tau_1 \xtimes \ldots \xtimes \tau_k)} }\\
\\
\tyrule{split}
  {\Gamma \vdash  e : \ctyp{\aclrd{\textbf{r}}\atimes\aclrd{\textbf{s}}}{(\tau_1 \xtimes \ldots \xtimes \tau_k)} }
  {\Gamma \vdash  \ident{split}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}}~e : \ctyp{\aclrd{\textbf{r}}}{(\tau_1 \xtimes \ldots \xtimes \tau_l)} \times \ctyp{\aclrd{\textbf{s}}}{(\tau_{l+1} \xtimes \ldots \xtimes \tau_k)}} \\
\\
\tyrule{dup}
  {\Gamma \vdash  e : \ctyp{\alift{\cclrd{r}\cpar\cclrd{s}}}{\tau} }
  {\Gamma \vdash  \ident{dup}_{\cclrd{r},\cclrd{s}}~e : \ctyp{\alift{\cclrd{r},\cclrd{s}}}{(\tau \xtimes \tau)} } \\
\end{array}
\end{equation*}

@@

\begin{theorem}[Well-typedness of the translation]
\label{thm:struct-welltyped}
Given a typing derivation for a well-typed closed expression $\coctx{}{\alift{}} \vdash e : \tau$
written in a structural context-aware programming langugae that is translated to the target language as
(we write $\ldots$ for the omitted part of the translation tree):
%
\begin{equation*}
  \semdeff
    {~~(\ldots)~~}
    {\coctx{}{\alift{}} \vdash e : \tau }
    {(\ldots)}
    { ~~f }
\end{equation*}
%
Then $f$ is well-typed, \ie~in the target language: $\vdash f : \sem{\coctx{()}{\alift{}}} \rightarrow \sem{\tau}$.
\end{theorem}
\begin{proof}
By rule induction over the derivation of the translation. Given a judgement
$\coctx{x_1\!:\!\tau_1 \ldots x_n\!:\!\tau_n}{\aclrd{\textbf{c}}} \vdash e : \tau$ where
$\aclrd{\textbf{c}} = \alift{\cclrd{c_1},\ldots,\cclrd{c_n}}$, the translation
constructs a function of type $\ctyp{\aclrd{\textbf{c}}}(\sem{\tau_1}\xtimes\ldots\xtimes\sem{\tau_n}}) \rightarrow \sem{\tau}$.

@@

\vspace{0.25em}\noindent\hangindent=0.6cm
Case (\emph{abs}): The type of $\ctx$ is $\ctyp{\aclrd{\textbf{r}}}(\ldots)$ and the type of $v$ is $\ctyp{\alift{\cclrd{s}}}\tau_1$,
  calling $\ident{merge}_{\aclrd{\textbf{r}},\alift{\cclrd{s}}}$ produces a context of type
  $\ctyp{\aclrd{\textbf{r}}\atimes\alift{\cclrd{s}}}(\ldots\xtimes\tau_1)$ as expected by $f$.

\vspace{0.25em}\noindent\hangindent=0.6cm
Case (\emph{app}): After applying $\ident{split}_{\aclrd{\textbf{r}}, \cclrd{t}\aseq\aclrd{\textbf{s}}}$ the types of
  $\ctx_1, \ctx_2$ are $\ctyp{\aclrd{\textbf{r}}}(\ldots)$ and $\ctyp{\cclrd{t}\aseq\aclrd{\textbf{s}}}(\ldots)$, respectively.
  $g$ requires $\ctyp{\aclrd{\textbf{s}}}(\ldots)$ and so the result of $\ident{cobind}_{\cclrd{t},\aclrd{\textbf{s}}}$ is
  $\ctyp{\alift{\cclrd{t}}}\tau_1$ as required by $f$.

\vspace{0.25em}\noindent\hangindent=0.6cm
Case (\emph{weak}): After applying $\ident{split}_{\aclrd{\textbf{r}}, \alift{\czero}}$ the type of
  $\ctx_1$ is $\ctyp{\aclrd{\textbf{r}}}(\ldots)$ as required.

\vspace{0.25em}\noindent\hangindent=0.6cm
Case (\emph{exch}), (\emph{contr}): The auxiliary definition $\ident{nest}_{\aclrd{\textbf{r}}, \aclrd{\textbf{s}}, \aclrd{\textbf{s'}}, \aclrd{\textbf{t}}}$
  keeps parts of the context corresponding to coeffect annotations $\aclrd{\textbf{r}}, \aclrd{\textbf{q}}$ unchanged and transforms
  the nested part. In (\emph{exch}), the provided lambda function is of type $\ctyp{\alift{\cclrd{s},\cclrd{t}}}{(\tau_1\xtimes\tau_2)}
  \rightarrow \ctyp{\alift{\cclrd{t},\cclrd{s}}}{(\tau_2\xtimes\tau_1)}$. In (\emph{contr}), the type of
  $\ident{dup}_{\cclrd{s},\cclrd{t}}$ is $\ctyp{\alift{\cclrd{s}\cpar\cclrd{t}}}{\tau} \rightarrow \ctyp{\alift{\cclrd{s},\cclrd{t}}}{(\tau\xtimes\tau)}$
  (assuming \ident{nest} is expanded, rather than treated as a value within the language).
\end{proof}

@@

\subsection{Structural coeffect language for dataflow}
\label{sec:struct-transl-safety}

The target language with comonadically-inspired primitives provides a framework that can be used
to model a variety of structural context-aware langauges and prove their safety. As outlined in
Section~\ref{sec:semantics-generalising}, the key principle that guarantees the safety of the
target language is a correspondence between coeffect annotations and the values they represent.
In a more expressive target language (such as Haskell or Agda), it would be sufficient to
provide an implementation of the comonadically-inspired primitives for the concrete domain-specific
language.

Our simple target language is not expressive enough to capture the correspondence in the type
system and so we instead follow the same methodology as when discussing safety for flat coeffect
languages in section Section~\ref{sec:semantics-proofs} and show safety for a sample concrete
context-aware language. We consider the structural coeffect language for dataflow. The definitions
can be compared with the flat version discussed in Section~\ref{sec:semantics-proofs-df}, which
highlights the similarities and differences between the flat and structural notion of context.

\begin{figure}[t]
\vspace{-0.5em}
\paragraph{Language syntax}
\begin{equation*}
\begin{array}{rcl}
v &=& \ldots \sep \ident{Df}\langle \textbf{v}_0, \ldots, \textbf{v}_k \rangle
    \hspace{9.5em} \textbf{v} ~=~ \langle v_0, \ldots, v_k \rangle \\
e &=& \ldots \sep \ident{Df}\langle \textbf{e}_0, \ldots, \textbf{e}_k \rangle \sep \ident{prev}_{\cclrd{n_0}, \ldots, \cclrd{n_k}}~e
    \quad\qquad \textbf{e} ~=~ \langle e_0, \ldots, e_k \rangle \\
K &=& \ldots \sep \ident{Df}\langle \textbf{v}_0, \ldots, \langle v_{j,0}, \ldots, v_{j, i-1}, \_, e_{j, i+1} \ldots, e_{j, n}\rangle, \ldots, \textbf{e}_k\rangle \\
   && \ldots \sep \ident{prev}_{\cclrd{n_0}, \ldots, \cclrd{n_k}}~\_
\end{array}
\end{equation*}

\vspace{0.75em}
\paragraph{Typing rules}

\begin{equation*}
\tyrule{vec}
  { \forall i\in\{0\ldots n\}. ~ \Gamma \vdash e_i : \tau}
  { \Gamma \vdash_{\ident{vec}} \langle e_0, \ldots, e_n \rangle : \tau, \cclrd{n} }
\end{equation*}
\begin{equation*}
\tyrule{df}
  { \forall i\in\{0\ldots k\}. ~ \Gamma \vdash_{\ident{vec}} \textbf{e}_i : \tau_i, \cclrd{n_i}}
  { \Gamma \vdash \ident{Df}\langle \textbf{e}_0, \ldots, \textbf{e}_k \rangle : \ctyp{\alift{\cclrd{n_0}, \ldots, \cclrd{n_k}}}{(\tau_0\xtimes\ldots\xtimes\tau_k)} }
\end{equation*}
\begin{equation*}
\tyrule{prev}
  { \Gamma \vdash e : \ctyp{\alift{\cclrd{n_0+1},\ldots,\cclrd{n_k+1}}}{(\tau_0\xtimes\ldots\xtimes\tau_k)} }
  { \Gamma \vdash \ident{prev}_{\cclrd{n_0},\ldots,\cclrd{n_k}}~e : \ctyp{\alift{\cclrd{n_0},\ldots,\cclrd{n_k}}}{(\tau_0\xtimes\ldots\xtimes\tau_k)} }
\end{equation*}

\vspace{1em}
\paragraph{Translation}

\begin{equation*}
\semdeff
  { \coctx{\Gamma}{\alift{\cclrd{n_0+1},\ldots,\cclrd{n_k+1}}} \vdash e : \tau }
  { \coctx{\Gamma}{\alift{\cclrd{n_0},\ldots,\cclrd{n_k}}} \vdash \kvd{prev}~e : \tau }
  {f}
  {\lambda ctx.\ident{prev}_{\cclrd{n_0},\ldots,\cclrd{n_k}}~\ctx }
\end{equation*}

\vspace{1em}
\paragraph{Reduction rules}

\begin{equation*}
\begin{array}{rl}
{\footnotesize(\emph{counit})} & \begin{array}{l}
\ident{counit}_{\cclrd{0}}(\ident{Df} \langle\langle v_0 \rangle\rangle)   \rightsquigarrow v_0
\end{array}
\\[0.25em]
{\footnotesize(\emph{cobind})} & \begin{array}{l}
\ident{cobind}_{\cclrd{m}, \alift{\cclrd{n_1},\ldots,\cclrd{n_k}}}~f~\ident{Df}\langle \langle a_{1,0}, \ldots a_{1,m+{n_1}} \rangle, \ldots, \langle a_{k,0}, \ldots a_{k,m+{n_k}} \rangle \rangle \rightsquigarrow \\[-0.25em]
\quad \ident{Df}\langle\langle\; f (\ident{Df}\langle \langle a_{1,0}, \ldots a_{1,{n_1}} \rangle, \ldots, \langle a_{k,0}, \ldots a_{k,{n_k}} \rangle \rangle), ~\ldots~, \\[-0.25em]
\qquad ~~~\, f (\ident{Df}\langle \langle a_{1,m}, \ldots a_{1,m+{n_1}} \rangle, \ldots, \langle a_{k,m}, \ldots a_{k,m+{n_k}} \rangle \rangle)~\; \rangle\rangle
\end{array}
\\[0.25em]
{\footnotesize(\emph{merge})} & \begin{array}{l}
\ident{merge}_{\alift{\cclrd{m_0},\ldots,\cclrd{m_k}}, \alift{\cclrd{n_0},\ldots,\cclrd{n_l}}} ((\ident{Df}\langle \textbf{v}_0, \ldots, \textbf{v}_k \rangle), (\ident{Df}\langle \textbf{v}'_0, \ldots, \textbf{v}'_l\rangle)) \rightsquigarrow
\\[-0.25em]
\quad \ident{Df}\langle \textbf{v}_0, \ldots, \textbf{v}_k, \textbf{v}'_0, \ldots, \textbf{v}'_l \rangle
\end{array}
\\[0.25em]
{\footnotesize(\emph{split})} & \begin{array}{l}
\ident{split}_{\alift{\cclrd{m_0},\ldots,\cclrd{m_k}}, \alift{\cclrd{n_0},\ldots,\cclrd{n_l}}} (\ident{Df}\langle \textbf{v}_0, \ldots, \textbf{v}_k, \textbf{v}'_0, \ldots, \textbf{v}'_l \rangle)  \rightsquigarrow
\\[-0.25em]
\quad (\ident{Df}\langle \textbf{v}_0, \ldots, \textbf{v}_k \rangle, ~\ident{Df}\langle \textbf{v}'_0, \ldots, \textbf{v}'_l\rangle)
\end{array}
\\[0.25em]
{\footnotesize(\emph{prev})} & \begin{array}{l}
\ident{prev}_{\alift{\cclrd{n_1},\ldots,\cclrd{n_k}}} (\ident{Df}\langle\langle v_{1,0}, \ldots, v_{1, n_1}, v_{1,{n_1}+1} \rangle,\ldots,\langle v_{k,0}, \ldots, v_{k, n_k}, v_{k,{n_k}+1} \rangle\rangle) \rightsquigarrow
\\[-0.25em]
\quad \ident{Df}\langle\langle v_{1,0}, \ldots, v_{1, n_1} \rangle,\ldots,\langle v_{k,0}, \ldots, v_{k, n_k} \rangle\rangle
\end{array}
\\[0.25em]
{\footnotesize(\emph{dup})} & \begin{array}{l}
\ident{dup}_{\cclrd{m},\cclrd{n}} (\ident{Df}\langle\langle v_0, \ldots, v_{\mathit{max}(m,n)} \rangle\rangle) \rightsquigarrow
  \ident{Df}\langle\langle v_0, \ldots, v_m \rangle, \langle v_0, \ldots, v_n \rangle\rangle
\end{array}
\end{array}
\end{equation*}

\caption{Additional constructs for modelling structural dataflow}
\label{fig:structural-ext-df}
\end{figure}

\paragraph{Domain-specific extensions.}

The Figure~\ref{fig:structural-ext-df} extends the target functional langauge with constructs,
typing rules, translation rules and reduction rules needed for modelling structural dataflow.
In the structural model, the type $\ctyp{\alift{\cclrd{r_0}, \ldots, \cclrd{r_n}}}{(\tau_0\xtimes\ldots\xtimes\tau_n)}$
represents a structure that provides values and additional contexts for variables of types
$\tau_0,\ldots,\tau_n$ with contextual capabilities as specified by corresponding coeffect annotations
$\cclrd{r_0}, \ldots, \cclrd{r_n}$.

In case of dataflow, the comonadically-inspired structure keeps vectors of past values for each of
the variables in the context. In the syntax, we write $\textbf{v}$ and $\textbf{e}$ for vectors of
values and expressions, respectively. The expression $\ident{Df}\langle \textbf{e}_0, \ldots, \textbf{e}_k \rangle$
is then formed by a vector of variable assignments where each variable assignment is a vector
of current and past values. When reducing expressions to values, we reduce values from left to
right and from current-most value to the last past value. This is specified by the context $K$.

When type-checking expressions that create the \ident{Df} values, we use an auxiliary judgement
$\Gamma \vdash_{\ident{vec}} e : \tau, \cclrd{n}$. The judgement checks that a vector of expressions
$\textbf{e}$ contains exactly $\cclrd{n}$ expressions of type $\tau$. This is caputred by the
(\emph{vec}) rule, which is then used to check individual elements of the context in the
(\emph{df}) rule.

Now consider a target language consisting of the core (ML-subset) defined by the syntax,
reduction rules and typing rules given in Figure~\ref{fig:semantics-target} (Chapter~\ref{ch:semantics})
with primitives inspired by structural indexed comonads defined in Figure~\ref{fig:struct-semantics-ext}
and also concrete notion of comonadically-inspired value and reduction rules for dataflow as
defined in Figure~\ref{fig:structural-ext-df}.

As with the examples discussed in Chapter~\ref{ch:semantics}, the resulting language is type safe.
Together with the well-typedness of the translation (Theorem~\ref{thm:struct-welltyped}), this
guarantees type safety of the structural coeffect calculus for dataflow.
In order to prove type safety, we first extend the \emph{canonical forms lemma}
(Lemma~\ref{thm:semantics-fp-canon}) and the \emph{preservation under substitution lemma}
(Lemma~\ref{thm:semantics-fp-pres-subst}). Those need to consider the new (\emph{df})
and (\emph{prev}) typing rules and substitution under the newly introduced expression forms
$\ident{Df}\langle\ldots\rangle$ and $\ident{prev}_{\cclrd{n}}$. We show that the translation
rule for \kvd{prev} produces well-typed expressions. Finally, we extend the type preservation
(Theorem~\ref{thm:structural-fp-pres}) and progress (Theorem~\ref{thm:structural-fp-prog}) theorems.

\begin{theorem}[Well-typedness of the \kvd{prev} translation]
\label{thm:structural-df-welltyped}
Given a typing derivation for a well-typed closed expression $\coctx{}{\alift{}} \vdash e : \tau$,
the translated program $f$ obtained using the rules in Figure~\ref{fig:struct-translation}
and Figure~\ref{fig:structural-ext-df} is well-typed, \ie~in the target language: $\vdash f : \sem{\coctx{()}{\alift{}}} \rightarrow \sem{\tau}$.
\end{theorem}
\begin{proof}
By rule induction over the derivation of the translation.

\vspace{0.5em}\noindent\hangindent=0.6cm
Case (\emph{var, num, abs, app}): As before.

\vspace{0.5em}\noindent\hangindent=0.6cm
Case (\emph{prev}): Type of $\ctx$ is $\ctyp{\alift{\cclrd{n_0+1},\ldots,\cclrd{n_k+1}}}{(\tau_0\xtimes\ldots\xtimes\tau_k)}$
  and so we can apply the (\emph{prev}) rule to obtain $\ctyp{\alift{\cclrd{n_0},\ldots,\cclrd{n_k}}}{(\tau_0\xtimes\ldots\xtimes\tau_k)}$
  as required by $f$.
\end{proof}

\begin{lemma}[Canonical forms]
\label{thm:structural-df-canon}
For all $e, \tau$, if $\vdash e : \tau$ and $e$ is a value then:
\begin{enumerate}
  \item If $\tau=\ident{num}$ then $e = n$ for some $n \in \mathbb{Z}$
  \item If $\tau=\tau_1 \rightarrow \tau_2$ then $e = \lambda x.e'$ for some $x, e'$
  \item If $\tau=\tau_1\times\ldots\times\tau_n$ then $e = (v_1, \ldots, v_n)$ for some $v_i$
  \item If $\tau=\ctyp{\alift{\cclrd{n_1}, \ldots, \cclrd{n_k}}}{(\tau_1\xtimes\ldots\xtimes\tau_k)}$
    then $e = \ident{Df}\langle \textbf{v}_0, \ldots \textbf{v}_n\rangle$ for some $\textbf{v}_i$
    such that $\textbf{v}_i = \langle v_0, \ldots, v_{n_i}\rangle$.
\end{enumerate}
\end{lemma}
\begin{proof}
  (1,2,3) as before; for (4) the last typing rule must have been (\emph{df}).
\end{proof}

\begin{lemma}[Preservation under substitution]
\label{thm:structural-df-pres-subst}
For all $\Gamma, e, e', \tau, \tau'$, if $\Gamma, x:\tau \vdash e : \tau'$ and $\Gamma \vdash e' : \tau$
then $\Gamma \vdash e[x \leftarrow e'] : \tau$.
\end{lemma}
\begin{proof}
  By induction over the derivation of $\Gamma, x:\tau \vdash e : \tau'$ as before, with new
  cases for $\ident{Df}\langle\ldots\rangle$ and $\ident{prev}_{\aclrd{\textbf{n}}}$.
\end{proof}

\begin{theorem}[Type preservation]
\label{thm:structural-df-pres}
  If $\Gamma \vdash e : \tau$ and $e \rightsquigarrow e'$ then $\Gamma \vdash e' : \tau$
\end{theorem}
\begin{proof}
  Rule induction over $\rightsquigarrow$.

\vspace{0.25em}\noindent\hangindent=0.6cm
Case (\emph{fn}, \emph{prj}, \emph{ctx}): As before, using Lemma~\ref{thm:structural-df-pres-subst} for (\emph{fn}).

\vspace{0.25em}\noindent\hangindent=0.6cm
Case (\emph{counit}): $e = \ident{counit}_{\cclrd{0}}(\ident{Df} \langle\langle v_0 \rangle\rangle)$. The last rule in the type
  derivation of $e$ must have been (\emph{counit}) with $\Gamma \vdash \ident{Df} \langle\lange v_0 \rangle\rangle : \ctyp{\alift{\cclrd{0}}}\tau$
  and therefore $\Gamma \vdash v_0 : \tau$.

\vspace{0.25em}\noindent\hangindent=0.6cm
Case (\emph{cobind}): $e = \ident{cobind}_{\cclrd{m}, \alift{\cclrd{n_1},\ldots,\cclrd{n_k}}}~f~(\ident{Df}\langle \textbf{v}_0, \ldots \textbf{v}_{k} \rangle)$
  such that $\forall i\in\{1\ldots k\}$ $\textbf{v}_i = \langle v_0, \ldots, v_{n_i}\rangle$. The last rule in the type derivation of $e$ must have been
  (\emph{cobind}) with a type $\tau = \ctyp{\alift{\cclrd{m}}}\tau'$ and assumptions:
\begin{itemize}[noitemsep,leftmargin=1.5cm,topsep=0pt]
  \item[--] $\Gamma \vdash f : \ctyp{\alift{\cclrd{n_1},\ldots,\cclrd{n_k}}}{(\tau_1\xtimes\ldots\xtimes\tau_k)} \rightarrow \tau_2$ and
  \item[--] $\Gamma \vdash \ident{Df}\langle \textbf{v}_0, \ldots \textbf{v}_{k} \rangle : \ctyp{\alift{\cclrd{m}+\cclrd{n_1},\ldots,\cclrd{m}+\cclrd{n_k}}}{(\tau_1\xtimes\ldots\xtimes\tau_k)}$
\end{itemize}
%
\hspace{0.5cm} Using the (\emph{df}) rule, the reduced expression has a type $\ctyp{\alift{\cclrd{m}}}\tau'$.

\vspace{0.25em}\noindent\hangindent=0.6cm
Case (\emph{merge}, \emph{split}, \emph{next}): Similar. In all three cases, the last typing rule in the derivation of $e$
  guarantees that the context contains correct number of vectors and each vector contains a
  sufficient number of values of a correct type.
\end{proof}

\begin{theorem}[Progress]
\label{thm:structural-df-prog}
  If $\vdash e : \tau$ then either $e$ is a value or there exists $e'$ such that $e \rightsquigarrow e'$
\end{theorem}
\begin{proof}
  By rule induction over $\vdash$.

\vspace{0.25em}\noindent\hangindent=0.6cm
Case (\emph{num,abs,var,app,proj,tup}): As before, using the adapted canonical forms lemma
  (Lemma~\ref{thm:structural-df-canon}) for (\emph{app}) and (\emph{proj}).

\vspace{0.25em}\noindent\hangindent=0.6cm
Case (\emph{counit}): $e = {\ident{counit}_\cunit}~e_1$. If $e_1$ is not a value, it can be reduced
  using (\emph{ctx}) with context ${\ident{counit}_\cunit}~\_$, otherwise it is a value. From Lemma~\ref{thm:structural-df-canon},
  $e_1 = \ident{Df}\langle\langle v \rangle\rangle$ and so we can apply (\emph{counit}) reduction rule.

\vspace{0.25em}\noindent\hangindent=0.6cm
Case (\emph{cobind}): $e = \ident{cobind}_{\cclrd{m}, \alift{\cclrd{n_1},\ldots,\cclrd{n_k}}}~e_1~e_2$. If $e_1$ is not a value,
  reduce using (\emph{ctx}) with context ${\ident{cobind}_{\cclrd{m},\alift{\cclrd{n_1},\ldots,\cclrd{n_k}}}}~\_~e$. If $e_2$ is
  not a value reduce using (\emph{ctx}) with context ${\ident{cobind}_{\cclrd{m},\alift{\cclrd{n_1},\ldots,\cclrd{n_k}}}}~v~\_$.
  If both are values then we have
  $e_2 = \ident{Df}\langle \langle a_{1,0}, \ldots a_{1,m+{n_1}} \rangle, \ldots, \langle a_{k,0}, \ldots a_{k,m+{n_k}} \rangle \rangle$
  from Lemma~\ref{thm:structural-df-canon} and so we can apply the (\emph{cobind}) reduction.

\vspace{0.25em}\noindent\hangindent=0.6cm
Case (\emph{merge}): $e = \ident{merge}_{\alift{\cclrd{m_0},\ldots,\cclrd{m_k}}, \alift{\cclrd{n_0},\ldots,\cclrd{n_l}}} e_1$. If $e_1$ is not a value,
  reduce using (\emph{ctx}) with context $e = \ident{merge}_{\alift{\cclrd{m_0},\ldots,\cclrd{m_k}}, \alift{\cclrd{n_0},\ldots,\cclrd{n_l}}}~\_$. If $e_1$ is
  a value, it must be a pair of vectors $(\ident{Df}\langle \textbf{v}_0, \ldots, \textbf{v}_k \rangle, \ident{Df}\langle \textbf{v}'_0, \ldots, \textbf{v}'_l\rangle)$
  using Lemma~\ref{thm:structural-df-canon} and it can reduce using (\emph{merge}) reduction.

\vspace{0.25em}\noindent\hangindent=0.6cm
Case (\emph{df}): $e = \ident{Df}\langle \textbf{e}_0, \ldots, \textbf{e}_n\rangle$. If $\textbf{e}_i$ is not a value
  then reduce using (\emph{ctx}) with the context $\ident{Df}\langle \ldots \rangle$.
  Otherwise, $\textbf{e}_0, \ldots, \textbf{e}_n$ are values and so $\ident{Df}\langle \textbf{e}_0, \ldots, \textbf{e}_n\rangle$
  is also a value.

\vspace{0.25em}\noindent\hangindent=0.6cm
Case (\emph{split}, \emph{prev}): Similar. Either sub-expression is not a value, or the type
  guarantees that it is a stream with correct number of elements to enable the (\emph{split})
  or (\emph{prev}) reduction, respectively.
\end{proof}

\begin{theorem}[Safety of context-aware dataflow language]
  If $\Gamma \vdash e : \tau$ and $e \rightsquigarrow^{*} e'$ then either $e'$ is a value of type $\tau$ or
  there exists $e''$ such that $e' \rightsquigarrow e''$ and $\Gamma \vdash e'' : \tau$.
\end{theorem}
\begin{proof}
  Rule induction over $\rightsquigarrow^*$ using Theorem~\ref{thm:structural-df-pres} and Theorem~\ref{thm:structural-df-prog}.
\end{proof}

@@

This chapter completes the key development of this thesis -- the presentation of the coeffect
framework, consisting of two calculi capturing properties of context-aware computations introduced
in Chapter~\ref{ch:applications}. In Chapters~\ref{ch:flat} and \ref{ch:semantics}, we focused on
whole-context properties of computations and we developed \emph{flat coeffect calculus} to capture them.
This chapter develops \emph{structural coeffect calculus}, which captures \emph{per-variable}
contextual properties. The system provides a precise analysis of liveness and dataflow and allows
other interesting uses such as tracking of variable accesses based on bounded linear logic.

Following the structure of the previous two chapters, the structural coeffect calculus is parameterized
by a \emph{structural coeffect algebra}. The two definitions are similar -- both require operations
$\cseq$ and $\cpar$ that model sequential and pointwise composition, respectively. For flat
coeffects, we required $\czip$ to model context merging. For structural coeffects, we
instead use a vector (free monoid) with the $\atimes$ operation -- which serves a similar purpose
as $\czip$. In order to keep track of separate annotations for each variable, we use a system with
explicit structural rules (contraction, weakening and exchange) that manipulate the structure
of variables and the structure of annotations at the same time.

The structural coeffect calculus has desirable equational properties that are satisfied only by
certain flat coeffect calculi. In particular, we show that $\beta$-reduction and $\eta$-expansion
have the typing preservation property for any instance of the structural coeffect calculus. These
two strong properties are desirable for programming languages, but are often not satisfied
(\eg~by languages with effects).

Finally, we discuss the semantics of the structural coeffect calculus in in terms of \emph{structural
indexed comonads}. As in Chapter~\ref{ch:semantics}, we first use the categorical semantics
to unify the different notions of context (bounded reuse, dataflow and liveness) and then use
it as a basis for translation that turns well-typed programs written in the structural coeffect
calculus into well-typed programs of a simple functional language. We give concrete reduction
rules for the target language, modelling \emph{structural dataflow} and prove type safety
of the language.
